Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control

### 2023-08-17 Rmarkdown/paper.Rmd
revision to text

Orthophosphate can control lead release into tap water, but its benefits are difficult to quantify since lead concentrations are so plumbing-dependent. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified. The best sentinel homes, then, are often short-lived. Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. In this study, geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Here we describe an alternative: sentinel lead pipe racks supplied with drinking water directly from the distribution system. While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible. Sentinel pipe racks can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind. And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022]. As highlighted previously [@masters_inherent_2016], they allow complete control of the flow rate, flow pattern, and stagnation time without involving or endangering consumers.

@@

A pronounced decreasing trend in lead concentration followed the increase to 2 mg PO~4~ L^-1^ (Figure 1a, c). Specifically, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a). The higher dose appeared to protect against lead release; and while this effect was slow to manifest, doubling the orthophosphate dose decreased the geometric mean lead concentration, year-over-year, by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).

Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding location-specific smooth term (Figure 1a). One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. While this did not impact distributed water pH, it did lower aluminum in treated water (Figure S6). And a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles [@kvech_role_2001]. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999]. But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data are not rich enough to identify the true cause.

The decrease in the location-specific trend representing Zone 2, however, followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b). Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b). Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%). Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the effect we attributed to a change in coagulation pH. That is, the hierarchical nature of the model allows us to control statistically for an effect common to all groups to better estimate an effect that occurred in only one group [@pedersen_hierarchical_2019].

@@

Even without an informative prior on $\beta$ (equation 2), the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks. This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016]. And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we should expect lower variability [@masters_inherent_2016].

@@

A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b). We used the sentinel pipe racks to determine the orthophosphate dose response in this zone. We estimated the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and the effect of a subsequent decrease to 1.5 mg PO~4~ L^-1^. But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016]. Seasonality was more complex in Zone 2, perhaps due to the inverse seasonality in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]). And since the dose increases occurred in late November and July in Zones 1 and 2 respectively, controlling explicitly for seasonality in the full model---encompassing both zones---was less important.

Temperature variation in the pipe rack effluent was a reasonable match for distributed water. While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked those of distributed water and point-of-use flushed samples more closely than the above-ground site (Rack 2), both captured the seasonal pattern well.

@@

Even after accounting for the seasonal variation in lead release, though, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a) and a 95% credible interval on the slope of the global trend that excluded zero. The intermediate dose appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses. Six months after the orthophosphate dose reduction, the increase in geometric mean lead release was estimated at `r round(pcent_increase$y[1])`%, with a 95% credible interval of `r round(pcent_increase$ymin[1])`--`r round(pcent_increase$ymax[1])`%.

>>>>

Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified. The best sentinel homes, then, are often short-lived. Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Here we describe an alternative: sentinel lead pipe racks supplied with drinking water directly from the distribution system. While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible. They can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind. And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022]. As highlighted previously [@masters_inherent_2016], they allow complete control of the flow rate, flow pattern, and stagnation time without involving or endangering consumers.

@@

A pronounced negative trend in lead concentration followed the increase to 2 mg PO~4~ L^-1^ (Figure 1a, c). Specifically, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a). The higher dose appeared to protect against lead release. And while this effect was slow to manifest, doubling the orthophosphate dose decreased the geometric mean lead concentration, year-over-year, by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).

Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding Zone 2-specific smooth term (Figure 1a). One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. While this did not impact distributed water pH, it did lower aluminum in treated water (Figure S6). And a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles [@kvech_role_2001]. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999]. But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data are not rich enough to identify the true cause.

The decrease in the location-specific trend representing Zone 2, however, followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b). Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b). Changes to the treatment process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%). Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the effect we attributed to a change in coagulation pH. That is, the hierarchical nature of the model allows us to control statistically for an effect common to all groups to better estimate an effect that occurred in only one group [@pedersen_hierarchical_2019].

@@

Even without an informative prior on $\beta$ (equation 2), however, the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks. This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016]. And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we expect lower variability [@masters_inherent_2016].

@@

A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b). We used the sentinel pipe racks to determine the orthophosphate dose response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^. But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016]. Seasonality was more complex in Zone 2, perhaps due to the inverse seasonal pattern in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]). And because the dose increase effect was estimated year-over-year in the full model (encompassing both zones) controlling for seasonality was less important.

Temperature variation in the pipe rack effluent was a reasonable match for distributed water. While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked distributed water and point-of-use flushed sample temperatures more closely than the above-ground site (Rack 2), both captured the seasonal pattern well. This, of course, is not necessarily true---in general, a comparison of pipe rack effluent and distributed water temperatures will help to illustrate the limitations of the pipe rack data.

@@

And even after accounting for the seasonal variation in lead release, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a)---and a 95% credible interval on the slope of the global trend that excluded zero. The intermediate dose appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses. Six months after the orthophosphate dose reduction, the increase in geometric mean lead release was estimated at `r round(pcent_increase$y[1])`%, with a 95% credible interval of `r round(pcent_increase$ymin[1])`--`r round(pcent_increase$ymax[1])`%.
